### 2023-08-21 sections/1-introduction.tex
lots of changes

\section{Introduction}
\label{sec:introduction}

Often dubbed "the Big Data explosion", rapidly advancing technologies have enabled researchers to collect complex data on an unprecedented scale. 
For example, the Sloan Digital Sky Survey (SDSS) has collected data on over 500 million celestial objects~\cite{alam2015eleventh}. Or, as another 
example, the National Center for Biotechnology Information (NCBI) has collected over 2.25 million ribosomal DNA sequences~\cite{silva2014silva}. Finally, 
as a third example, the Large Hadron Collider (LHC) has collected over 100 petabytes of data~\cite{evans2008lhc}. Or, as a fourth example, the 
National Oceanic and Atmospheric Administration (NOAA) has collected over 20 petabytes of data~\cite{noaa2019noaa}. 
In many cases, the influx of data outpaces the rate of improvement in computing capacity predicted by Moore's Law~\cite{moore1965cramming}, indicating 
that computer hardware will not "catch up" to our computational needs in the near future. As a result, analyzing and learning from these large datasets 
requires better algorithms.

@@

Naive implementations of kNN, whose asymptotic complexity is a function of the dataset's cardinality, 
prove prohibitively slow for large datasets. While algorithms for kNN on large datasets exist, most of them are approximate and fail to 
exploit the inherent structure and redundancy often present in big datasets. While approximate search is sufficient for many applications, the 
need for efficient, exact search remains.

Recent approaches to tackle the exponential growth of data include locality-sensitive hashing~\cite{bingham2001random}, 
clever indexing techniques such as the FM Index~\cite{ferragina2005indexing}, and entropy-scaling search~\cite{yu2015entropy}. 
Entropy-scaling search is a paradigm for similarity search that exploits the inherent structure and redundancy often present in big datasets.
Importantly, as suggested by their name, entropy-scaling search algorithms have asymptotic complexity that scales with geometric properties of the dataset,
rather than its cardinality. In 2019, we introduced CHESS (CLustered Hierarchical Entropy-Scaling Search), which extended entropy-scaling $\rho$-nearest 
neighbors search to a hierarchical clustering approach. In this paper, we introduce CLAM (Clustered Learning of Approximate Manifolds), and CAKES (CLAM-Accelerated K-nearest-neighbor 
Entropy-scaling Search). CLAM, like CHESS, is a hierarchical clustering algorithm (figure out and explain improvements upon CHESS present in CLAM if there are any algorithmic ones)
that can facilitate a variety of algorithms for big data. Using the tree constructed by CLAM, CAKES builds upon the capabilities of CHESS to perform 
k-nearest neighbors search in addition to $\rho$-nearest neighbors search.

At some point, need to address the fact that exact search is important for small k and approxiate 
is maybe more ideal for large k. Try to find citation for this.

Growth in size and types of datasets:
\begin{itemize}
    \item SDSS APOGEE2 data. Near-infrared spectra of approximately 430,000 stars in approximately 9,000 dimensions. $R=22,500 = \frac{\lambda}{\Delta \lambda}$ for $\lambda \in [1.51, 1.70) \mu m$
    \item SDSS MaNGA hexagon shaped images. 10,000 galaxies with 2,700 deg$^2$, resolution $R=2,000$ for $\lambda \in [360, 1000) nm$
    \item Silva 18S ribosomal DNA sequences of approximately 2.25 million genomes with aligned length of 50,000 letters. TODO: Look into other variants in the dataset particularly NR99.
    \item Datasets from ANN-Benchmarks suite
\end{itemize}

Recent approaches in dealing with exponential growth of data:
\begin{itemize}
    \item LSH
    \item clever indexing e.g. FM Index.
    \item Entropy-scaling search~\cite{yu2015entropy} and CHESS~\cite{ishaq2019clustered}.
\end{itemize}

>>>>

\section{Introduction}
\label{sec:introduction}
Often dubbed "the Big Data explosion", rapidly advancing technologies have enabled researchers to collect complex data on an unprecedented scale. 
The Sloan Digital Sky Survey (SDSS), for example, has collected near-infrared spectra of approximately 430,000 stars in approximately 9,000 dimensions~\cite{alam2015eleventh}. 
Add sentence about MaNGA here: 10,000 galaxies with 2,700 deg$^2$, resolution $R=2,000$ for $\lambda \in [360, 1000) nm$.

Biology has also seen an influx of large, high-dimensional datasets. Silva 18S contains ribosomal DNA sequences of approximately 2.25 
million genomes with aligned length of 50,000 letters. (Do we need to look at NR99??)

In many cases, the increase in data outpaces the rate of improvement in computing capacity predicted by Moore's Law~\cite{moore1965cramming}, indicating 
that computer hardware will not "catch up" to our computational needs in the near future. As a result, analyzing and learning from these large datasets 
requires better algorithms.

@@

Naive implementations of kNN, whose asymptotic complexity is a function of the dataset's cardinality, 
prove prohibitively slow for large datasets. While algorithms for kNN on large datasets exist, most of them are approximate and fail to 
exploit the inherent structure and redundancy often present in big datasets. Approximate search is sufficient for many applications, but the 
need for efficient, exact search remains.

Recent approaches to tackling the exponential growth of data include locality-sensitive hashing~\cite{bingham2001random}, 
clever indexing techniques such as the FM Index~\cite{ferragina2005indexing}, and entropy-scaling search~\cite{yu2015entropy, ishaq2019clustered}. 
Entropy-scaling search is a paradigm for similarity search that exploits the inherent structure and redundancy often present in big datasets.
Importantly, as suggested by their name, entropy-scaling search algorithms have asymptotic complexity that scales with geometric properties of the dataset,
rather than its cardinality. In 2019, we introduced CHESS (CLustered Hierarchical Entropy-Scaling Search), which extended entropy-scaling $\rho$-nearest 
neighbors search to a hierarchical clustering approach. In this paper, we introduce CLAM (Clustered Learning of Approximate Manifolds), and CAKES (CLAM-Accelerated K-nearest-neighbor 
Entropy-scaling Search). CLAM, like CHESS, is a hierarchical clustering algorithm (figure out and explain improvements upon CHESS present in CLAM if there are any algorithmic ones)
that can facilitate a variety of algorithms for big data. Using the tree constructed by CLAM, CAKES builds upon the capabilities of CHESS to perform 
k-nearest neighbors search in addition to $\rho$-nearest neighbors search.

At some point, need to address the fact that exact search is important for small k and approximate 
is maybe more ideal for large k. Try to find citation for this.

### 2023-08-21 sections/2-methods.tex
lots of changes

\subsection{Partition}
\label{subsec:methods:partition}

Partition algorithm for building the tree.
Lift this from the CHAODA paper.

Exact partition costing $\mathcal{O}(n^2)$ vs approximate partition using $\sqrt{n}$ seeds to achieve $\mathcal{O}(n)$ cost.

% latex does not seem to like math symbols in this heading
\subsection{\texorpdfstring{$\rho$}{p}-Nearest Neighbors Search}
\label{subsec:methods:rnn-search}

Modified binary search.
Lift this from CHESS paper.

\subsection{k-Nearest Neighbors Search}
\label{subsec:methods:knn-search}

Repeated $\rho$-NN search with where radius is modified with each iteration as guided by lfd of query ball.

Lift from CHESS paper as a starting point.

\subsection{Compression}
\label{subsec:methods:compression}

Encode and Decode algorithms.
These will be specific for each dataset and/or distance function.

% discrete vs continuous, and small alphabet vs large alphabet in discrete

>>>>

In this manuscript, we are primarily concerned with $k$-nearest neighbors search in a finite-dimensional 
vector space. Some of the algorithms we present use a modified version of the $\rho$-nearest neighbors search 
algorithm described in CHESS~\cite{ishaq2019clustered} to conduct $k$-NN search. (TODO: make second section less 
vague once we determine exactly what we're doing with sharding and indexing.)

Given a dataset $\textbf{X} = \{x_1 \dots x_n\}$, we define a \emph{point} or \emph{datum} $x_i \in \textbf{X}$ as a singular observation (e.g., the genome of 
an organism, the vector representation of a single image, or any entity on which we can define a \emph{distance function}).

We define a \emph{distance function} $f : (\textbf{X}, \textbf{X}) \mapsto \mathbb{R}^+ \cup \{0\}$ as a function which, 
given two points $x, y \in \textbf{X}$, deterministically returns a non-negative real number. We require that the distance function 
be symmetric ($f(x, y) = f(y, x) \forall x, y \in \textbf{X}$) and that the distance between two points $x$ and $y$ be zero if and only if $x = y$. 
When, in addition to these constraints, that the distance function obeys
the triangle inequality, it is a \emph{distance metric}, and in such cases we can guarantee that our search algorithms have perfect recall. 
Choice of distance function varies by type of data. For example, with electromagnetic spectra, we use both 
Euclidean (L2) and Cosine distances. With biological or string data on the other hand, Hamming and Levenshtein distances are more appropriate.

With these concepts defined, we can now pose the $k$- and $\rho$ nearest neighbors search problems.
Given a query $q$, along with a distance function $f$ defined on a dataset $ \textbf{X}$, $k$-nearest neighbors search aims to: 
for a given $k$, find the $k$ closest points to $q$ in $ \textbf{X}$; that is, find the $k$ points $x \in \textbf{X}$ with the smallest $f(q, x)$.
We also have the $\rho$-nearest neighbors search problem: for a given radius $\rho$, find all points in $\textbf{X}$ that are a distance of $\rho$ 
or less from $q$; that is, all points $x \in \textbf{X}$ such that $f(q, x) \leq \rho$.

Some of our algorithms for $\rho$- and $k$-nearest neighbors search rely on the \emph{local fractal dimension} at some region of the dataset, 
which we define as: 
$$\text{log}_2 (\frac{|B_D(q, r_1)|}{|B_D(q, r_2)|})$$
where $B_D(q, r)$ is the set of points contained in a ball of radius $r$ 
centered on a point $q$ in the dataset $D$; in this example, we compute fractal dimension for some radius $r_1$ and a smaller radius $r_2 = \frac{r_1}{2}$.
We stress that this concept of local fractal dimension differs from the \emph{embedding dimension} of a dataset. To illustrate the difference,
consider SDSS's APOGEE dataset, wherein each datum is a nonnegative real-valued vector of length 8,575. Hence, the \emph{embedding dimension} of this dataset is 8,575. 
However, due to physical constraints (namely, the laws that govern stellar fusion and spectral emission lines), the data are constrained to a lower-dimensional 
manifold. The \emph{local fractal dimension} is an approximation of the dimensionality of that lower-dimensional manifold at a given point, for some length scale.
The notion that high-dimensional data collected from constrained generating phenomena typically only occupy a low-dimensional manifold is known as the \emph{manifold hypothesis}.

We define a \emph{cluster} as a set of points with a \emph{center}, a \emph{radius}, and an approximation of the \emph{local fractal dimension}.
The \emph{center} is the geometric median of the set of points in the \emph{cluster}, and so it is a real data point. The \emph{radius} is the
maximum distance from a point in the cluster to its \emph{center}. We estimate \emph{local fractal dimension} at the cluster radius and half
the cluster radius. Each cluster (unless it is a leaf cluster) has two child clusters in much the same way that a node in
a binary tree has two child nodes. We define the \emph{metric entropy} of a data set under a hierarchical clustering scheme as a refinement of [7], where
metric entropy for a given cluster radius $r_c$ was the number of clusters of radius $r_c$ needed to cover all data. Here, we use a binary, divisive clustering 
approach, but with early stopping criteria; clusters with fewer than a threshold number of points are not further split. 
Since we frame the asymptotic complexity of $\rho$-NN search in terms of the number of leaf clusters, the \emph{metric entropy} is best thought of in terms of the number of
leaf clusters.

\subsection{Clustering}
\label{subsec:methods:clustering}

We start by building a divisive hierarchical clustering of the dataset with CLAM, using a 
similar recursive procedure as outlined in CHESS~\cite{ishaq2019clustered}, but with the following 
improvements: better selection of poles for partitioning and memoization of useful cluster properties, 
such as local fractal dimension (see Section ADD SECTION: $k$-NN by Repeated $\rho$-NN).

CLAM assumes the manifold hypothesis. 
In other words, we assume that the dataset is embedded in a $D$-dimensional space, but that the data only occupy 
a $d$-dimensional manifold, where $d \ll D$. 
While we sometimes use Euclidean notions, such as voids and volumes, to talk about geometric and topological 
properties of clusters and of the manifold, CLAM does not rely on such notions; 
they serve merely as a convenient, more intuitive way to talk about the underlying mathematics.

For a cluster with $m$ points, we begin by computing a 
random sample of $\sqrt m$ points. We then compute the geometric median of this sample, which we call the 
cluster's center. We then compute the distance from the center to each point in the sample of $\sqrt m$ points. 
The point $l$ which is furthest from the center is designated the left pole, and the point $r$ which is furthest
from $l$ is designated the right pole. We then partition the cluster into a left child and a right child, where the 
left child contains all points in the cluster which are closer to $l$ than to $r$, and the right child contains all 
points in the cluster which are closer to $r$ than to $l$. Without loss of generality, we assign to the left child 
those points which are equidistant from $l$ and $r$. Starting from a root-cluster containing the entire dataset, we 
repeat this procedure until each leaf contains only one datum or until some other user-specified stopping criterion 
is met.

\begin{algorithm} % enter the algorithm environment
\caption{Partition} % give the algorithm a caption
\label{alg:partition} % and a label for \ref{} commands later in the document
\begin{algorithmic}[1] % enter the algorithmic environment
    \REQUIRE $cluster$
    \STATE $m \leftarrow \lfloor \sqrt{|cluster.points|} \rfloor$
    \STATE $seeds \leftarrow m$ random points from $cluster.points$
    \STATE $c \leftarrow$ geometric median of $seeds$
    \STATE $l \leftarrow \argmax d(c,x) \ \forall \ x \in cluster.points$
    \STATE $r \leftarrow \argmax d(l,x) \ \forall \ x \in cluster.points$
    \STATE $left \leftarrow \{x | x \in cluster.points \land d(l,x) \le d(r,x)\}$
    \STATE $right \leftarrow \{x | x \in cluster.points \land d(r,x) < d(l,x)\}$
    \IF{$|left| > 1$}
        \STATE Partition($left$)
    \ENDIF
    \IF{$|right| > 1$}
        \STATE Partition($right$)
    \ENDIF
\end{algorithmic}
\end{algorithm}

\subsubsection {Guaranteed Decrease in Cluster Radii}
\label{subsubsec:methods:guaranteed-decrease-in-cluster-radii}

We show that cluster radius is guaranteed to decrease after at most $d$ partitions.

Assume that we have a dataset which follows a $d$-dimensional distribution embedded in $\mathcal{D}$-dimensional space.
We can describe this distribution choosing some set of $d$ mutually orthogonal axes.
Let $m$ denote the maximum distance between any pair of points along any axis from any choice of mutually orthogonal axes for our $d$-dimensional distribution.
In the worst case scenario, the distribution of our data is a $d$-sphere and $m$ is the maximum distance between two points along every axis for every choice of axes.
In this scenario, the cluster radius is $r = \frac{m}{2}$.

Given a cluster $\mathcal{C}$, our partition algorithm will choose the left and right poles, 
and the axis defined by those poles will be the axis for partitioning $\mathcal{C}$'s points into two child clusters.
For points in each child cluster, the maximum distance between any pair of points along the \emph{axis of partitioning} will have been reduced to at most $\frac{m}{2}$.
Though this bound holds for points on the axis of partitioning, it does not serve as a bound for the maximum distance between \emph{any} pair of points in $\mathcal{C}$.
To see this, consider the two dimensional scenario where the points in $\mathcal{C}$ are distributed in a circle of radius $r$. Clearly, the maximum possible distance between 
any two points in $\mathcal{C}$ is $2r$, and occurs when the points fall on opposite ends of one of its diameters. Now consider a horizontal diameter of the circle, 
and let $l$ denote the leftmost point on this diameter. Next, consider the orthogonal vertical diameter, and let $x$ denote the uppermost point on this diameter. When we 
partition along the horizontal diameter, $l$ and $x$ will be assigned to the left child (as mentioned in ADD SECTION, we assign equidistant points to the left child). 
This reduces the maximum possible distance between any two points on the axis of partition -- the horizontal diameter -- to $r$. However, by considering the 
right triangle formed by $l$, $x$, and the center of the circle, we can see that the distance between 
$l$ and $x$ is $\sqrt{2}{r}$.

If $d > 1$ then the child clusters will have a maximal chord $m_{child}$ such that $m_{child} \leq m$.
By partitioning the cluster in this way, we have taken one axis and reduced the maximum distance along that axis from $m$ to $\frac{m}{2}$.
However, as per the distribution of data as discussed earlier, there may be other axes along which the maximum pairwise distance is still $m$.
We, thus, recursively partition each child cluster.

With each recursive application of Partition, we consume one axis along which the maximum distance was $m$ and reduce the maximum distance along that axis to $\frac{m}{2}$ in the child clusters.
After $d$ recursive applications of partition, we will have exhausted the $d$ axes with the maximum distance of $m$.
After these $d$ partitions, the points in each child cluster will have a maximum pairwise distance of $\frac{m}{2}$ along any axis.
Thus the radii of those child clusters will be at most $\frac{m}{4}$.

\subsubsection {Complexity}
\label{subsubsec:methods:clustering:complexity}

Clustering: Exact partition with $\mathcal{O}(n^2)$ cost vs approximate partition using $\sqrt{n}$ seeds to achieve $\mathcal{O}(n)$ cost.
Building the exact tree costs $\mathcal{O}(n^2 \log n)$ vs approximate tree for $\mathcal{O}(n \log n)$.

\subsection {Index-Building}
\begin{itemize}
    \item sharded search 
    \item index building 
    \item optimization of number of shards 
\end{itemize}

\subsection{\texorpdfstring{$\rho$}{p}-Nearest Neighbors Search}
\label{subsec:methods:rnn-search}

Given a query $q$ and a search radius $\rho$, find all $x \in X$ s.t. $f(q, x) \leq \rho$.

We conduct ranged-nearest neighbors search similarly as described in CHESS~\cite{ishaq2019clustered}, but 
with some implementation improvements including identification of candidate clusters which are leaves and 
conducting exhaustive search over those leaves.

The asymptotic complexity of $\rho$-nearest neighbors is the same as in CHESS~\cite{ishaq2019clustered}.

\subsection{\texorpdfstring{$k$}{k}-Nearest Neighbors Search}
\label{subsec:methods:knn-search}

The $k$-nearest neighbors search problem is posed as follows: given a query $q$ and a positive integer $k$, find the $k$ closest instances in $X$.
We present five variants of $k$-nearest neighbors search: Greedy Search, K-NN by Repeated $\rho$-NN, Sieve Search I, Sieve Search II, and 
Linear Search. We use a preliminary auto-tuning function to predict the variant which will perform 
best for a given query, dataset, and value of $k$. We then proceed with search using that variant.

Given a Cluster $C$, let $c$ be its center and $r$ be its radius. Our $k$-NN variants make use of the following cluster 
properties:
\begin{itemize}
    \item $\delta = f(q, c)$ is the distance from the query to the cluster center $c$.
    \item $\delta_{max} = \delta + r$ is the distance from the query to the potentially farthest instance in $C$.
    \item $\delta_{min} = \text{max}(0, \delta - r)$ is the distance from the query to the potentially closest instance in $C$.
\end{itemize}

We sometimes also use the notation $\delta = f(x, q)$ with points instead of clusters to refer to the distance from a point $x$ to the query $q$.

\subsubsection{Greedy Search}
\label{subsubsec:methods:knn-search:greedy-search}
Greedy Search repeatedly partitions the cluster with the lowest $\delta_{min}$ into its children 
until it is a leaf, removing it and adding its points to a fixed size priority queue of points. Search terminates 
when the furthest hit in the queue of points has $\delta$ greater than the $\delta_{min}$ of the closest remaining
cluster.

Formally, let $Q$ be a priority queue of clusters sorted by non-increasing $\delta_{min}$. Initialize $Q$ with the root cluster.
Let $P$ be a second priority queue of points sorted by non-decreasing distance from the query. This queue starts empty.

While $P$ has fewer than $k$ clusters or $\delta$ of the top priority element in $P$ is greater 
than $\delta_{min}$ of the top priority element in $Q$, do the following:
\begin{enumerate}
\item While the top priority element $q$ in $Q$ isn't a leaf, replace $q$ with its children.
\item Remove $q$ from $Q$ and add all of its points to $P$. 
\item Remove points from $P$ until $P$ has $k$ points. 
\end{enumerate}

Return $P$.

\begin{algorithm} 
\caption{Greedy Search} 
\label{alg:greedy_search} 
\begin{algorithmic}[1]
    \REQUIRE $tree$, $query$, $k$
    \STATE $Q \leftarrow$ priority queue
    \STATE $Q.push(tree.root)$
    \STATE $P \leftarrow$ priority queue
    \WHILE{$|P| < k \lor P.max.\delta > Q.min.\delta_{min}$}
        \WHILE{$!Q.min.isLeaf$}
            \STATE $[l, r] \leftarrow Q.extractMin().children()$
            \STATE $Q.push([l, r])$
        \ENDWHILE
        \STATE $leaf \leftarrow Q.extractMin()$
        \STATE $P.push(leaf)$
        \WHILE{$|P| > k$}
            \STATE $P.extractMax()$
        \ENDWHILE
    \ENDWHILE
    \STATE Return $P$
\end{algorithmic}
\end{algorithm}

\subsubsection{$k$-NN by Repeated $\rho$-NN}
\label{subsubsec:methods:knn-search:repeated-rnn}

Hence its name, $k$-NN by Repeated $\rho$-NN repeatedly performs 
$\rho$-nearest neighbors search, increasing the radius until $k$ neighbors
have been found.

Search starts with a radius $m$ equal to the radius of the cluster tree divided by
the cardinality of the dataset. We perform $\rho$-NN search with radius $m$. 
If no points are within a distance $m$ of the query, we increase the radius by a factor of 
2 and perform $rho$-NN search again, repeating until at least one point is found.

Once we have found at least one point, we continue to perform $\rho$-NN search, but instead of 
increasing the radius by a factor of 2 on each iteration, we increase it by a factor determined 
by the local fractal dimension of the clusters containing the points found so far. In particular, 
we increase the radius by a factor of $\text{max}(2, \frac{k}{|P|^{\frac{1}{\mu}}}$), where $|P|$ is the number of 
points found so far and $\mu$ is the mean local fractal dimension of the clusters containing the points found so far.
Since the local fractal dimension is memoized at clustering time, this poses no additional cost at search time.
Once $|P| >= k$, we sort the points in $P$ by $\delta$ and return the $k$ closest points.

TODO: 
- add intutiion for factor 
- introduce P variable sooner

\subsubsection{Sieve Search I}
\label{subsubsec:methods:knn-search:sieve}

\subsubsection{Sieve Search II}

In this variant of KNN, we repeatedly calculate a threshold distance $\tau$ such that no cluster with a $\delta_{min}$ greater than $\tau$ 
can contain one of the $k$ nearest neighbors. Search terminates when we have a threshold containing exactly $k$ points.

Let $Q$ be a vector of clusters sorted by non-decreasing $\delta_{max}$. Initialize $Q$ with  
the root cluster. 
Let $P$ be a priority queue of size $k$ containing points sorted by non-decreasing distance from the query. This queue starts empty.

Repeat the following: 
\begin{enumerate}
\item Find the cluster $C_{\tau}$ in $Q$ with the smallest $\delta_{max}$ such that no cluster whose $\delta_{min}$ is greater than $C_{\tau}$'s $\delta_{max}$ can contain one of the $k$ nearest neighbors.
\item Let the threshold $\tau$ be the $\delta_{max}$ of $C_{\tau}$.
\item Remove from $P$ any points which are a distance greater than $\tau$ from the query. 
\item If no cluster in $Q$ has a $\delta_{max}$ less than or equal to $\tau$, break.
\item Remove from $Q$ any cluster whose $\delta_{min}$ is greater than $\tau$.
\item Transfer from $Q$ to $P$ all the points in any cluster whose cardinality is less than $k$ or who is a leaf. 
\item Replace all non-leaf clusters in $Q$ with their children. 
\end{enumerate}

At this stage, if any clusters remain in $Q$, we add all their points to $P$. 
Extract $P$.

\begin{algorithm} % enter the algorithm environment
    \caption{Sieve V1} % give the algorithm a caption
    \label{alg:sieve_v1} % and a label for \ref{} commands later in the document
    \begin{algorithmic}[2] % enter the algorithmic environment
        \REQUIRE $tree$, $query$, $k$
        \STATE $Q \leftarrow$ priority queue
        \STATE $Q.push(tree.root)$
        \STATE $P \leftarrow$ priority queue
        \WHILE{$|P| < k \lor P.max.\delta > Q.max.\delta_{min}$}
            \WHILE{$!Q.extractMax().isLeaf$}
                \STATE $[l, r] \leftarrow Q.extractMax().children()$
                \STATE $Q.push([l, r])$
            \ENDWHILE
            \STATE $leaf \leftarrow Q.extractMax()$
            \STATE $P.push(leaf)$
            \WHILE{$|P| > k$}
                \STATE $P.extractMax()$
            \ENDWHILE
        \ENDWHILE
        \STATE Return $P$
    \end{algorithmic}
    \end{algorithm}

\subsubsection{Linear Search}
\label{subsubsec:methods:knn-search:leaf-search}

Let $hits$ be a a fixed size priority queue with $k$ instances sorted by non-decreasing $f(q, x)$, i.e. distance to the query.
Initialize $hits$ with the centers of $candidates$.

Let $f_k$ be the distance to the $k^{th}$ farthest instance so far.
Exclude all $candidates$ for which $\delta^2 > f_k$, i.e. their closest possible instance is farthest that the $k^{th}$ farthest instance found so far.
Remove the closest cluster from $candidates$ and add all its contained instances to $hits$.
Repeat until $candidates$ is empty.

\subsubsection{Complexity}
\label{subsubsec:methods:knn-search:complexity}

This is at-least on-par with the complexity for $\rho$-nearest neighbors search from CHESS.
It is potentially better because we effectively shrink the search radius to adapt to the $k$ closest instances found during the algorithm.

\label{subsubsec:methods:clustering:proof-of-eventually-decreasing-cluster-radii}

\subsubsection {Complexity}

\subsection{\texorpdfstring{$\rho$}{p}-Nearest Neighbors Search}
\label{subsec:methods:rnn-search}

Given a query $q$ and a search radius $\rho$, find all $x \in X$ s.t. $f(q, x) \leq \rho$.

Modified binary search to identify candidate leaf clusters.
Exhaustive leaf search over those leaves.

Lift this from CHESS paper. Be sure to cite CHESS here.

Complexity is the same as in CHESS.

\subsection{\texorpdfstring{$k$}{k}-Nearest Neighbors Search}
\label{subsec:methods:knn-search}

The k-nearest neighbors search problem is posed as follows: given a query $q$ and a positive integer $k$, find the $k$ closest instances in $X$.

Given a Cluster $C$, let $c$ be its center and $r$ be its radius. Our $k$-nn variants make use of the following definitions:
\begin{itemize}
    \item $\delta = f(q, c)$ is the distance from the query to the cluster center.
    \item $\delta_{max} = \delta + r$ is the distance from the query to the potentially farthest instance in the Cluster.
    \item $\delta_{min} = \text{max}(0, \delta - r)$ is the distance from the query to the potentially closest instance in the Cluster.
\end{itemize}

\subsubsection{Expanding Threshold}
\label{subsubsec:methods:knn-search:expanding-threshold}
In this variant of KNN, we repeatedly

Let $Q$ be a priority queue of clusters sorted by non-increasing $\delta_{min}$. Initialize $Q$ with the root cluster.
Let $P$ be a second priority queue of points sorted by non-decreasing distance from the query. This queue starts empty.

While $P$ has fewer than $k$ clusters or $\delta$ of the top priority element in $P$ is greater 
than $\delta_{min}$ of the top priority element in $Q$, do the following:
\begin{enumerate}
\item While the top priority element $q$ in $Q$ isn't a leaf, replace $q$ with its children.
\item Remove $q$ from $Q$ and add all of its points to $P$. 
\item Remove points from $P$ until $P$ has $k$ points. 
\end{enumerate}

Return $P$.

\begin{algorithm} % enter the algorithm environment
\caption{Expanding Threshold} % give the algorithm a caption
\label{alg:expanding_threshold} % and a label for \ref{} commands later in the document
\begin{algorithmic}[1] % enter the algorithmic environment
    \REQUIRE $tree$, $query$, $k$
    \STATE $Q \leftarrow$ priority queue
    \STATE $Q.push(tree.root)$
    \STATE $P \leftarrow$ priority queue
    \WHILE{$|P| < k \lor P.max.\delta > Q.min.\delta_{min}$}
        \WHILE{$!Q.min.isLeaf$}
            \STATE $[l, r] \leftarrow Q.extractMin().children()$
            \STATE $Q.push([l, r])$
        \ENDWHILE
        \STATE $leaf \leftarrow Q.extractMin()$
        \STATE $P.push(leaf)$
        \WHILE{$|P| > k$}
            \STATE $P.extractMax()$
        \ENDWHILE
    \ENDWHILE
    \STATE Return $P$
\end{algorithmic}
\end{algorithm}

\subsubsection{Sieve V1}
\label{subsubsec:methods:knn-search:sieve-v1}

In this variant of KNN, we repeatedly calculate a threshold distance $\tau$ such that no cluster with a $\delta_{min}$ greater than $\tau$ 
can contain one of the $k$ nearest neighbors. Search terminates when we have a threshold containing exactly $k$ points.

Let $Q$ be a vector of clusters sorted by non-decreasing $\delta_{max}$. Initialize $Q$ with  
the root cluster. 
Let $P$ be a priority queue of size $k$ containing points sorted by non-decreasing distance from the query. This queue starts empty.

Repeat the following: 
\begin{enumerate}
\item Find the cluster $C_{\tau}$ in $Q$ with the smallest $\delta_{max}$ such that no cluster whose $\delta_{min}$ is greater than $C_{\tau}$'s $\delta_{max}$ can contain one of the $k$ nearest neighbors.
\item Let the threshold $\tau$ be the $\delta_{max}$ of $C_{\tau}$.
\item Remove from $P$ any points which are a distance greater than $\tau$ from the query. 
\item If no cluster in $Q$ has a $\delta_{max}$ less than or equal to $\tau$, break.
\item Remove from $Q$ any cluster whose $\delta_{min}$ is greater than $\tau$.
\item Transfer from $Q$ to $P$ all the points in any cluster whose cardinality is less than $k$ or who is a leaf. 
\item Replace all non-leaf clusters in $Q$ with their children. 
\end{enumerate}

At this stage, if any clusters remain in $Q$, we add all their points to $P$. 
Extract $P$.

\begin{algorithm} % enter the algorithm environment
    \caption{Sieve V1} % give the algorithm a caption
    \label{alg:sieve_v1} % and a label for \ref{} commands later in the document
    \begin{algorithmic}[2] % enter the algorithmic environment
        \REQUIRE $tree$, $query$, $k$
        \STATE $Q \leftarrow$ priority queue
        \STATE $Q.push(tree.root)$
        \STATE $P \leftarrow$ priority queue
        \WHILE{$|P| < k \lor P.max.\delta > Q.max.\delta_{min}$}
            \WHILE{$!Q.extractMax().isLeaf$}
                \STATE $[l, r] \leftarrow Q.extractMax().children()$
                \STATE $Q.push([l, r])$
            \ENDWHILE
            \STATE $leaf \leftarrow Q.extractMax()$
            \STATE $P.push(leaf)$
            \WHILE{$|P| > k$}
                \STATE $P.extractMax()$
            \ENDWHILE
        \ENDWHILE
        \STATE Return $P$
    \end{algorithmic}
    \end{algorithm}

\subsubsection{Sieve V2}
\label{subsubsec:methods:knn-search:sieve-v2}
Add explanation of the sieve v2 algorithm.

\subsubsection{Repeated Rnn}
\label{subsubsec:methods:knn-search:repeated-rnn}
Add explanation of the repeated rnn algorithm.

\subsubsection{Leaf Search}
\label{subsubsec:methods:knn-search:leaf-search}

Let $hits$ be a a fixed size priority queue with $k$ instances sorted by non-decreasing $f(q, x)$, i.e. distance to the query.
Initialize $hits$ with the centers of $candidates$.

Let $f_k$ be the distance to the $k^{th}$ farthest instance so far.
Exclude all $candidates$ for which $\delta^2 > f_k$, i.e. their closest possible instance is farthest that the $k^{th}$ farthest instance found so far.
Remove the closest cluster from $candidates$ and add all its contained instances to $hits$.
Repeat until $candidates$ is empty.

\subsubsection{Complexity}
\label{subsubsec:methods:knn-search:complexity}

Clustering: Exact partition with $\mathcal{O}(n^2)$ cost vs approximate partition using $\sqrt{n}$ seeds to achieve $\mathcal{O}(n)$ cost.
Building the exact tree costs $\mathcal{O}(n^2 \log n)$ vs approximate tree for $\mathcal{O}(n \log n)$.

This is at-least on-par with the complexity for $\rho$-nearest neighbors search from CHESS.
It is potentially better because we effectively shrink the search radius to adapt to the $k$ closest instances found during the algorithm.

### 2023-08-21 sections/4-results.tex
lots of changes

For each pair $(X, f)$ and a small selection of depths, measure:
\begin{enumerate}[1.]
    \item linear search vs accelerated search:
    \begin{enumerate}[i.]
        \item runtime performance,
        \item speedup factor, and
        \item number of distance comparisons.
    \end{enumerate}
    \item Compression CAKES + zlib vs just zlib, also substitute other libraries for zlib:
    \begin{enumerate}[i.]
        \item just the encoding vs the full dataset,
        \item encoding + serialization, and
        \item encoding + serialization + zlib.
    \end{enumerate}
    \item Measure time taken for accelerated-search in the compressed space and decompression of the hits with different combinations of serde and codec libraries.
\end{enumerate}

Find the optimal clusters for compression.
Compare linear search vs accelerated search terminated at these optimal clusters.

>>>>

For each pair $(X, f)$ and a small selection of depths, measure:
\begin{enumerate}[1.]
    \item linear search vs accelerated search:
    \begin{enumerate}[i.]
        \item runtime performance,
        \item speedup factor, and
        \item number of distance comparisons.
    \end{enumerate}
\end{enumerate}
